---
title: "Getting started with anatomogramdata"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with anatomogramdata}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 5,
  fig.height = 6.5
)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
knitr::opts_chunk$set(eval = has_ggplot2)
```

```{r setup}
library(anatomogramdata)
library(ggplot2)
```

anatomogramdata is a **data package, not a plotting package**: it ships
tidy, ontology-tagged human anatomogram polygon data (`hgMale`/`hgFemale`),
parsed directly from the [EBI Expression Atlas anatomogram
source](https://github.com/ebi-gene-expression-group/anatomogram) — not a
frozen hardcoded snapshot. There is no custom geom; you plot the bundled
data with `ggplot2::geom_polygon()` or `ggplot2::geom_map()` directly,
which already provide full aesthetic flexibility.

`anatomogram_select()` is the package's single entry point for browsing,
selecting, and filtering that data. It has three mutually exclusive modes.

## Browse mode

Call it with no arguments to see what's available — every tissue and
which organ system(s) it belongs to:

```{r}
head(anatomogram_select("male"))
unique(anatomogram_select("male")$system)
```

## Organ mode: select specific organs, optionally with values

Pass `organs` as a character vector. Each element is matched either as a
literal ontology ID (e.g. `"UBERON_0000948"`) or fuzzily (case-insensitive,
partial) against the tissue name. Name an element to attach a value to it
— useful for `fill`:

```{r}
d <- anatomogram_select("male", organs = c(kidney = "High", heart = "Low"))

ggplot(d, aes(x, y, group = group)) +
  geom_polygon(data = subset(d, tissue_id == "outline"),
               fill = "grey90", colour = "black") +
  geom_polygon(data = subset(d, tissue_id != "outline"),
               aes(fill = value), colour = "black") +
  coord_fixed() +
  theme_void()
```

Unnamed elements are pure selection, with no value attached:

```{r}
selection_only <- anatomogram_select("male", organs = c("kidney", "heart"))
```

An organ name that matches zero tissues warns and is dropped rather than
failing the whole call; a name that matches more than one distinct tissue
(e.g. `"hippocampus"`, which has a left and a right entry under separate
ontology IDs) is an error listing the candidates, since the ambiguity
needs a decision you have to make, not one the package should guess at.

## System mode: pull whole organ systems

Pass `system` instead of `organs` to select every tissue belonging to one
or more organ systems (Skeletal, Digestive, Nervous, and so on — see
[organ_systems] for the full list). The outline is included automatically:

```{r}
bones <- anatomogram_select("male", system = "Skeletal")

ggplot(bones, aes(x, y, group = group)) +
  geom_polygon(data = subset(bones, tissue_id == "outline"),
               fill = "grey95", colour = "black") +
  geom_polygon(data = subset(bones, tissue_id != "outline"),
               fill = "grey40", colour = "black") +
  coord_fixed() +
  theme_void()
```

Requesting more than one system is facet-ready — a tissue belonging to
more than one requested system (the pancreas, under both Digestive and
Endocrine) correctly appears once per system:

```{r}
several <- anatomogram_select("male", system = c("Digestive", "Endocrine"))

ggplot(several, aes(x, y, group = group)) +
  geom_polygon(data = subset(several, tissue_id == "outline"),
               fill = "grey95", colour = "black") +
  geom_polygon(data = subset(several, tissue_id != "outline"),
               fill = "steelblue", colour = "black") +
  facet_wrap(~ system) +
  coord_fixed() +
  theme_void()
```

`organs` and `system` are mutually exclusive in a single call — specifying
both is an error. Call `anatomogram_select()` twice and `rbind()` the
results if you genuinely need both.

## The outline: silhouette vs. interior contour lines

`hgMale`/`hgFemale`'s `outline_role` column splits the body outline into
three roles, computed from geometry rather than any tag in the source SVG
(there isn't one): `"silhouette"` (the plain outer body shape),
`"contour_detail"` (the same shape with interior finger/toe/joint lines),
and `"feature"` (small facial marks). Filter to just the roles you want:

```{r}
outline <- hgMale[hgMale$tissue_id == "outline", ]
clean_silhouette <- outline[outline$outline_role == "silhouette", ]

ggplot(clean_silhouette, aes(x, y, group = group)) +
  geom_polygon(fill = "grey90", colour = "black") +
  coord_fixed() +
  theme_void()
```

## Power-user plotting with `ggplot2::geom_map()`

`anatomogram_select()` covers the common cases, but `hgMale`/`hgFemale`
already carry the `id`/`group` columns and pre-flipped `y` that
`ggplot2::geom_map()` needs as its `map` argument directly — useful when
mapping more than one aesthetic to real data (e.g. `fill` *and*
`linewidth` together). See `examples/geom_map_demo.R` and
`examples/cause_of_death_demo.R` in the package source for worked
patterns.

## Attribution

The bundled data is parsed directly from the EBI Expression Atlas
anatomogram source (Apache-2.0 code / CC-BY-4.0 images) — see
`citation("anatomogramdata")` for how to cite both the package and the
underlying source.
